---
title: "Reproducible Research: Peer Assessment 1"
subtitle: "Coursera project"
date: "May, 2019"
author: "David Affagard"
output: 
  html_document:
    keep_md: true
---
[Git repo : https://github.com/affagard/RepData_PeerAssessment1/](https://github.com/affagard/RepData_PeerAssessment1/), forked from [rdpeng](https://github.com/rdpeng/RepData_PeerAssessment1)


\  

\  

# Libraries required for this report
```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(lubridate)
library(stringr)
library(knitr)
```

\  

\  

## Loading and preprocessing the data
- Data source URL : https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip [52K]
- Dataset : Data about personal movement using activity monitoring devices. The device collects data at 5 minutes interval through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November 2012. It includes the number of steps taken in 5 minutes interval each day.
\  

#### Init values
```{r}
        data_folder <- "data"
        fig_folder <- "figure"
        data_zip_name <- "activity.zip"
        data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```
\  

#### Dowload data source if not already done
```{r}
        if(!file.exists(data_zip_name)) {
                
                download.file(data_url, destfile = data_zip_name, method = "curl")
                
        }
```
\  

#### Unzip data source if not already done
```{r}
        if(!dir.exists(data_folder)) {
                
                dir.create(data_folder, showWarnings = TRUE, recursive = FALSE, mode = "0744")
                unzip(data_zip_name, exdir = data_folder)
                
        }
```
\  

#### Create a local folder for figures
```{r}
        if(!dir.exists(fig_folder)) {
                
                dir.create(fig_folder, showWarnings = TRUE, recursive = FALSE, mode = "0744")
                
        }
```
#### Load activity dataset, (`activity`)
```{r}
        activity <- read.csv (paste(data_folder,"/activity.csv", sep=""), header=TRUE, sep = ",", dec=".")
```
\  

#### Tidy data : Column (`activity$date`) as date
```{r}
        activity$date <- as.Date(activity$date)
```
\  

#### Tidy data : Format column (`activity$interval`) as (`%H%M`) character pattern
```{r}
        activity$interval <- str_pad(as.character(activity$interval), 4, pad = "0")
```      


\  

\  

## What is mean total number of steps taken per day?
#### Calculate total number of steps taken each day (`nb_steps_each_day`)
```{r}
        # Group by date and summarise per day
        nb_steps_each_day <- group_by(activity, date)
        # Summarise and set column names
        nb_steps_each_day <- summarise(nb_steps_each_day, steps_a_day = sum(steps, na.rm = TRUE))
        names(nb_steps_each_day) <- c("day", "steps_a_day")
```
\  

#### Mean and Median to draw their vertical lines
```{r}
        # Mean
        mean_total_nb_steps <- mean(nb_steps_each_day$steps_a_day, na.rm = TRUE)
        # Median
        median_total_nb_steps <- median(nb_steps_each_day$steps_a_day, na.rm = TRUE)
```
\  

#### Histogram for **Total number of steps taken each day** (`nb_steps_each_day`)

```{r}
        g <- ggplot(
                data = nb_steps_each_day,
                aes(x = steps_a_day)
        )
        g <- g + theme_light()
        g <- g + geom_histogram(bins = 30, color = "#666666", fill = "white", show.legend = FALSE, na.rm = FALSE)
        g <- g + ggtitle("Total number of steps taken each day (figure/plot1.png)")
        g <- g + xlab("Steps / day")
        # Mean line
        g <- g + geom_vline(
                        xintercept = mean_total_nb_steps,
                        linetype = "dashed",
                        color = "#ff6666"
                        )
        # Mean label
        g <- g + geom_label(
                        x = 3000,
                        y = 9.5,
                        hjust = "left",
                        label = paste(
                                "Mean = ",
                                round(mean_total_nb_steps,0),
                                " steps",
                                sep=""
                                ),
                        color="#ff6666"
                        )
        # Median line
        g <- g + geom_vline(
                        xintercept = median_total_nb_steps,
                        linetype = "dashed",
                        color = "#6666ff"
                        )
        # Median label
        g <- g + geom_label(
                        x = 12000,
                        y = 9.5,
                        hjust = "left",
                        label = paste(
                                "Median = ",
                                round(median_total_nb_steps,0),
                                " steps",
                                sep=""
                        ),
                        color="#6666ff"
                        )
        print(g)
```

- Mean total number of steps taken each day : **`r format(round(mean(nb_steps_each_day$steps_a_day),0), big.mark = " ")` steps **

- Median total number of steps taken each day : **`r format(round(median(nb_steps_each_day$steps_a_day),0), big.mark = " ")` steps**

\  

\  

[File : figure/plot1.png](figure/plot1.png)
```{r warning=FALSE}
        png(
                filename = paste(fig_folder,"plot1.png",sep="/"),
                width = 640,
                height = 640,
                units = "px",
                pointsize = 12,
                bg = "white"
        )
        print(g)
        dev.off()
                
```


\  

\  

## What is the average daily activity pattern?

The average daily activity pattern is figured with 

- A time series of the 5-minutes interval during a day on x-axis
- The average number of steps taken, averaged across all days on y-axis
- In follows, (`daily_activity`) is the data frame to get the daily activity pattern
\  

#### Group by interval and summarise by the mean
```{r}
        daily_activity <- group_by(activity, interval)
        daily_activity <- summarise(daily_activity, average_steps = mean(steps, na.rm = TRUE))
```
\  

#### Convert column interval into POSIXct
```{r}
        daily_activity$interval <- parse_date_time(daily_activity$interval, "HM") + months(9) + years(2012)
```           
\  

#### Plotting average daily activity

We use ggplot2 library

First, we calculate the 5-minute interval containing the maximum steps
```{r}
max_average <- max(daily_activity$average_steps)
interval_max_steps <- subset(daily_activity, average_steps == max_average)
interval_max_steps_ftime <- date_format("%Hh%M")(interval_max_steps$interval)
```

The maximum number of steps averaged across days (rounded `max_average`) is **`r round(max_average,0) ` steps**, it happens at **`r interval_max_steps_ftime`** (`interval_max_steps_ftime`)

\  

So we init a dash lines (`hline`) and (`vline`), and a text (`htext`) to point out in blue the maximum value of averages daily activity

```{r}
        hline <- geom_hline(
                yintercept = max_average,
                linetype = "dashed",
                color = "#6666ff"
        )

        vline <- geom_vline(
                xintercept = interval_max_steps$interval,
                linetype = "dashed",
                color = "#6666ff"
        )

        htext <- geom_label(
                x = interval_max_steps$interval,
                y = max_average,
                hjust = "left",
                vjust = "bottom",
                label = paste(
                        "Maximum of activity : ",
                        round(max_average,0),
                        " steps at ",
                        interval_max_steps_ftime,
                        sep=""
                ),
                color="#6666ff"
        )
        
```
\  

Then we set and print the ggplot instance, a png file is generated in folder figure

```{r warning=FALSE}
        g <- ggplot(
                data = daily_activity,
                aes(x = interval, y = average_steps, na.rm = TRUE)
        )
        g <- g + geom_line(aes(color = "red"), show.legend = FALSE) + theme_light()
        g <- g + ylim(c(0,max(daily_activity$average_steps)*1.1))
        g <- g + geom_smooth(method = "loess", color = "#6666ff", bg = "#ccccee")
        g <- g + ggtitle("Daily activity pattern, 5 minutes averaged over 2 months, (figure/plot2.png)")
        g <- g + ylab("Steps") + xlab("Time on a day")
        g <- g + scale_x_datetime(labels = date_format("%Hh%M"))
        g <- g + hline + htext + vline
        print(g)
```

[File : figure/plot2.png](figure/plot2.png)
```{r warning=FALSE}
        png(
                filename = paste(fig_folder,
                "plot2.png",sep="/"),
                width = 640,
                height = 640,
                units = "px",
                pointsize = 12,
                bg = "white"
        )
        print(g)
        dev.off()
```


\  

\  

## Imputing missing values

There is **`r format(sum(is.na(activity$steps), na.rm = FALSE), big.mark = " ")` missing values**, i.e. **`r round(sum(is.na(activity$steps), na.rm = FALSE)/nrow(activity)*100,2)`%** of the values, in the dataset `activity`
```{r}
        n_missing_values <- sum(is.na(activity$steps), na.rm = FALSE)
        print(n_missing_values)
```
\  

**Hypothesis and method to filled-in missing values :**

- We guess the best way to extrapolate missing steps (`NA` values) is to replace missing value by the average daily activity corresponding to its 5-minutes interval.

- We are alerted by the fact that 2 days have daily activity under 5% of the average daily steps, we could consider them as abnormal and decide to replace all of their interval steps values by their average daily activity. Nevertheless, since it's not required in assigments, we assume to not alter theses "surely-dummy" days.

- We set a new dataset (`activity_filled`) by replacing missing values of steps.

- Since (`NA`) steps data are replaced by decimal averages, we round them down to set a real case.
\  

\  

Replace the **`r format(sum(is.na(activity$steps), na.rm = FALSE), big.mark = " ")`** `NA`values
```{r}
        activity_filled <- activity
        for( k in rep(1:length(activity_filled$steps))){
                if(is.na(activity_filled$steps[[k]])){
                        activity_filled$steps[[k]] <- round(daily_activity$average_steps[[1+k%%288]],0)
                }
        }
        activity_filled$steps <- as.integer(activity_filled$steps)
```
\  

Check the total number of missing values is now 0 for the new dataset (`activity_filled`)
```{r}
        print(sum(is.na(activity_filled$steps), na.rm = FALSE))
```
\  


#### New Histogram for total number of steps taken each day, with filled-in missing values
Calculate the new total number of steps taken each day (`nb_steps_each_day_filled`), summarise it and name columns
```{r}
        # Group by date and summarise per day
        nb_steps_each_day_filled <- group_by(activity_filled, date)
        # Summarise by sum and set column names
        nb_steps_each_day_filled <- summarise(nb_steps_each_day_filled, steps_a_day = sum(steps, na.rm = TRUE))
        names(nb_steps_each_day_filled) <- c("day", "steps_a_day")
```
\  

Horizontal line to draw for mean and median 
```{r}
        # Mean
        mean_total_nb_steps_filled <- mean(nb_steps_each_day_filled$steps_a_day, na.rm = TRUE)
        
        # Median
        median_total_nb_steps_filled <- median(nb_steps_each_day_filled$steps_a_day, na.rm = TRUE)
```

```{r}
        g <- ggplot(
                data = nb_steps_each_day_filled,
                aes(x = steps_a_day)
        )
        g <- g + theme_light()
        g <- g + geom_histogram(bins = 30, color = "#666666", fill = "white", show.legend = FALSE, na.rm = FALSE)
        g <- g + ggtitle(
                "Steps taken each day, with filled-in missing values, (figure/plot3.png)"
        )
        g <- g + xlab("Steps / day")
        
        # Mean line
        g <- g + geom_vline(
                        xintercept = mean_total_nb_steps_filled,
                        linetype = "dashed",
                        color = "#ff6666"
                )
        
        # Mean label
        g <- g + geom_label(
                        x = 3000,
                        y = 11,
                        hjust = "left",
                        label = paste(
                                "Mean = ",
                                round(mean_total_nb_steps_filled,0),
                                " steps",
                                sep=""
                        ),
                        color="#ff6666"
                )
        
        # Median line
        g <- g + geom_vline(
                        xintercept = median_total_nb_steps_filled,
                        linetype = "dashed",
                        color = "#6666ff"
                )
        
        # Median label
        g <- g + geom_label(
                        x = 12000,
                        y = 11,
                        hjust = "left",
                        label = paste(
                                "Median = ",
                                round(median_total_nb_steps_filled,0),
                                " steps",
                                sep=""
                        ),
                        color="#6666ff"
                )
        
        print(g)
```

[File : figure/plot3.png](figure/plot3.png)
```{r warning=FALSE}
        png(
                filename = paste(fig_folder,
                "plot3.png",sep="/"),
                width = 640,
                height = 640,
                units = "px",
                pointsize = 12,
                bg = "white"
        )
        print(g)
        dev.off()
```

- Mean total number of steps taken each day : **`r format(round(mean(nb_steps_each_day_filled$steps_a_day),0), big.mark = " ")` steps **

- Median total number of steps taken each day : **`r format(round(median(nb_steps_each_day_filled$steps_a_day),0), big.mark = " ")` steps**

- On left, we see 2 days with full zero step, which are not processed by NA treatment

\  
With filled-in missing values, 

 - the mean is closer to the median,
 
 - the mean and the median are higher.


\  

### Are there differences in activity patterns between weekdays and weekends?
\  

#### Set a new dataset (`activity_w`) from (àctivity_filled`)
With a new field (`day_type`) as.factor with 2 levels "weekday", "weekend"
```{r}
        activity_w <- activity_filled
        # Add column day_type, init with "weekday"
        activity_w$day_type <- "weekday"
        
        # For weekend days, replace "weekday" by "weekend"
        activity_w$day_type[which(wday(activity_w$date, week_start = 1) %in% c(6,7))] <- "weekend"
```
\  

#### Daily activity pattern (`daily_activity_w`)
From (`activity_w`) grouped by (`day_type`) and (`ìnterval`) and summarised by the mean
```{r}
        daily_activity_w <- group_by(activity_w, interval, day_type)
        daily_activity_w <- summarise(daily_activity_w, average_steps = mean(steps, na.rm = TRUE))
```
\  

#### Convert column interval into POSIXct
```{r}
        daily_activity_w$interval <- parse_date_time(daily_activity_w$interval, "HM") + months(9) + years(2012)
```           
\  

#### Plotting average daily activity by comparing weekdays / weekends
```{r warning=FALSE}
        g <- ggplot(
                data = daily_activity_w,
                aes(x = interval, y = average_steps, color = day_type, group = day_type)
        )
        g <- g + geom_line(show.legend = TRUE) + theme_light()
        g <- g + ylim(c(0,max(daily_activity_w$average_steps)*1.05))
        g <- g + geom_smooth(method = "loess", color = "#6666ff", bg = "#ccccee")
        g <- g + ggtitle("Daily activity pattern, 5 minutes averaged over 2 months, (figure/plot4.png)")
        g <- g + ylab("Steps") + xlab("Time on a day")
        g <- g + scale_x_datetime(labels = date_format("%Hh%M")) + scale_color_discrete(name = "")
        g <- g + facet_grid(rows = vars(day_type))
        print(g)
```

[File : figure/plot4.png](figure/plot4.png)
```{r warning=FALSE}
                png(
                        filename = paste(fig_folder,"plot4.png",sep="/"),
                        width = 640,
                        height = 640,
                        units = "px",
                        pointsize = 12,
                        bg = "white"
                )
                print(g)
                dev.off()
                
```